---
# Become a Data Wizard with R, Chapter 10: companion file
title: "When do Atlantic storms form?"
author: "Your name"
date: today
format:
  html:
    embed-resources: true
  docx: default
execute:
  echo: false
---

```{r}
#| label: setup
#| message: false
library(dplyr)
library(readr)
library(ggplot2)

storms_raw <- read_csv("data/storms.csv")
```

## Question

In which months of the year do Atlantic storms form, and how many
of them become hurricanes? A storm here is a tropical or
subtropical cyclone whose winds reached at least 34 knots
(63 km/h), the strength at which storms receive a name.

## Data

The file `data/storms.csv` holds the `storms` data from the dplyr
package, built from the Atlantic hurricane database (HURDAT2) of
the US National Hurricane Center. It has
`r format(nrow(storms_raw), big.mark = ",")` rows, one for each
storm at each time it was measured. The analysis starts in 1979,
because the rows for earlier years are incomplete, and counts
each storm in the month of its first row as a cyclone.

```{r}
#| label: clean
# The text and the captions say 1979-2024; stop if the data differ
stopifnot(max(storms_raw$year) == 2024)

cyclone <- c("tropical depression", "tropical storm", "hurricane",
             "subtropical depression", "subtropical storm")

storm_list <- storms_raw |>
  filter(year >= 1979) |>
  # Zeta's January 2006 rows continue a storm that began in 2005
  filter_out(name == "Zeta", year == 2006) |>
  # Keep only the times when the system was a cyclone
  filter(status %in% cyclone) |>
  # Names come back every six years, so a storm is a name in a year
  group_by(name, year) |>
  summarize(
    month = min(month),
    peak_kt = max(wind),
    hurricane = any(status == "hurricane"),
    .groups = "drop"
  ) |>
  # Drop the depressions that never reached tropical-storm strength
  filter(peak_kt >= 34)

n_storms <- nrow(storm_list)
pct_aug_oct <- round(100 * mean(storm_list$month %in% 8:10))
pct_sep <- round(100 * mean(storm_list$month == 9))
pct_hurricane <- round(100 * mean(storm_list$hurricane))
```

## Findings

From 1979 to 2024, `r n_storms` Atlantic storms reached
tropical-storm strength. Of these, `r pct_aug_oct`% formed in
August, September, or October, and September alone accounted for
`r pct_sep`% (@fig-months). In all, `r pct_hurricane`%
became hurricanes, with winds of at least 64 knots (119 km/h);
@tbl-months gives the figures month by month. The months with
few storms, such as January and December, give percentages that
rest on a handful of cases.

```{r}
#| label: fig-months
#| fig-cap: "Atlantic storms by month of formation, 1979–2024."
#| fig-width: 5.6
#| fig-height: 3
outcome_colors <- c("Below hurricane strength" = "grey75",
                    "Hurricane" = "#1F4FA8")

storm_list |>
  mutate(
    month = factor(month.abb[month], levels = month.abb),
    outcome = if_else(hurricane, "Hurricane",
                      "Below hurricane strength")
  ) |>
  ggplot(aes(x = month, fill = outcome)) +
  geom_bar() +
  scale_x_discrete(drop = FALSE) +
  scale_fill_manual(values = outcome_colors) +
  labs(x = NULL, y = "Storms", fill = NULL,
       caption = "Data: NOAA HURDAT2, from the storms data in dplyr") +
  theme_minimal(base_size = 9) +
  theme(legend.position = "top",
        panel.grid.major.x = element_blank())
```

```{r}
#| label: tbl-months
#| tbl-cap: "Storms and hurricanes by month of formation, 1979–2024."
storm_list |>
  group_by(month) |>
  summarize(
    storms = n(),
    hurricanes = sum(hurricane),
    pct = round(100 * hurricanes / storms),
    peak_kmh = round(median(peak_kt * 1.852))
  ) |>
  mutate(month = month.abb[month]) |>
  knitr::kable(col.names = c("Month", "Storms", "Hurricanes",
                             "Hurricanes (%)",
                             "Median peak wind (km/h)"))
```

## Software

```{r}
#| label: software
sessionInfo()
```
